---
title: "Prep"
author: "Mike Malloy"
date: "5/16/2021"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)


req_pkg <- c("tidyverse", "httr", "janitor", "readxl", "tidycensus", "sf", "stringi", "leaflet")

check.and.install.Package<- function(package_name){
  
        if(!package_name %in% installed.packages()) {
           install.packages(package_name)
        }
}

for (i in req_pkg) {
  
  check.and.install.Package(i)
  
  library(i, character.only = TRUE)
  
}

```

## Read in Excel file with Fiscal Disparities from 2019
```{r}

## assign label to URL for the dataset
url <- "https://metrocouncil.org/Communities/Publications-And-Resources/Fiscal-Disparities-Summary-Data-2014.aspx"

## use GET function from httr package to retrieve URL
## will store as a temporary file; setting file extension ensures file is read in correctly
GET(url, write_disk(tf <- tempfile(fileext = ".xlsx")))

## Use read_excel to read file in as data frame
## After row 194, the sheet has summary tables with different column structure. Stop reading in data there (193 in code bc skipping row 1 entirely)
fd19 <- read_excel(tf, sheet = "FD summary data 2021", col_names = F, skip = 1, n_max = 193)

```



## Clean Fiscal Disparities file

Reformat and clean the column names
```{r}

## Need to clean the column names. Must collapse columns 1 through 4 into a single string
cols <- apply(fd19[1:4,], 2, paste, collapse = "")

cols <- gsub("NA", "", cols)

## two records have the page title cell combined, delete those portions
cols[4] <- "County"

cols[6] <- "Contribution Tax Base"

## rename the city field, we'll use that to join later and it's very long
cols[5] <- "Community"

## assign column names
colnames(fd19) <- cols

## simple clean to remove white space, make all lower case
fd19 <- clean_names(fd19)

## remove rows 1:4 now
fd19 <- fd19[5:193,]

```

Clean community names so they'll match later
```{r}

## A few of these say '(combined)' - we want to remove this

fd19 <- fd19 %>%
  mutate(community = gsub(" \\(combined\\)", "", community))

```

Make columns numeric
```{r}

fd19 <- fd19 %>%
  mutate(across(c(6:24), as.numeric))

```


## Calculate percentage change in CI tax base after FD transfer

```{r}

fd19 <- fd19 %>%
  mutate(pct_change_ci_base = round(net_change_tax_base_payable_2021/contribution_tax_base*100, 2))

```


```{r eval=FALSE, include=FALSE}

## Check  

fd19 %>%
  select(community,
         county,
         net_change_tax_base_payable_2021, 
         contribution_tax_base,
         distribution_tax_base_payable_2021,
         pct_change_ci_base) %>%
  View()

```


## Pull in ACS data on median household income for communities in the Twin Cities metro area. 

Using ACS 5-year estimates from 2018. Variables are found by searching the ACS API variables here: https://api.census.gov/data/2018/acs/acs5/profile/variables.html

```{r}

my_api_key = ("73f7f4579cc4b9b6ef927d785abf0dee32ccb86a")

tc_acs <- get_acs(geography = "county subdivision",
                  variables = c("DP03_0062E"),
                  state = "MN",
                  geometry = TRUE,
                  key = my_api_key)

```

## Clean ACS file

Split the name field into city/town/county/state columns so we can match 
```{r}

## just city/county/state at first, separated at comma
tc_acs <- tc_acs %>%
  separate(col = NAME, into = c("community", "county", "state"), sep = ",") %>%
  mutate(county = trimws(county)) %>%
  select(-state)

## now separate community name from the city/town/village descriptor
tc_acs <- tc_acs %>%
  mutate(community_name = substr(community, 1, stri_locate_last_fixed(community, " ")-1),
         tcv = substr(community, stri_locate_last_fixed(community, " ")+1, nchar(community))) %>%
  select(-community)

## If tcv value = "township", I want to add " Twp." to community name, so it'll match the fd19.
tc_acs <- tc_acs %>%
  mutate(community_name = if_else(tcv == "township", paste(community_name, "Twp."), community_name))

## Confirm I will be able to join on this field eventually

fd19 %>% filter(!community %in% tc_acs$community)

## They don't all match but the ones that don't make sense bc not census areas (and barely partaking in FD): airports, state fairgrounds, Fort Snelling.

```

Rename estimate field, just so it's more clear
```{r}

tc_acs <- tc_acs %>%
  rename(med_income = estimate)

```



## Combine into one base file

Join ACS onto the trimmed FD table
```{r}

base <- fd19 %>%
  mutate(county = paste(county, "County")) %>% ## so that it matches with ACS, which adds word "county" to each value
  inner_join(tc_acs, by = c("community" = "community_name", "county" = "county")) %>%
  st_sf() %>% ## makes dataframe an sf object which preserves geometry from ACS file; needed for mapping
  st_transform(crs = 4326)

```


## Prep for mapping

Examine variables to determine map scales

```{r}

summary(base$pct_change_ci_base)

boxplot(base$pct_change_ci_base)

## Alright so some negative, most positive but below 100, some way over 100
## Want to capture net loser/net gainer dynamic, and also highlight the real outliers.
## So I think best categories are <0, 0 - 100, and >100. 
## Will do a diverging gradient palette with more detail for first map.

summary(base$med_income)

boxplot(base$med_income)

## Pretty tight around median...

## Actually, could use federal poverty level. Have done that before. At like 200%, 400%, >400%.
## nope tried (fam of 4 - 52400, and 104800), too few below poverty level. Just do some neat numbers roughly around there?
tabyl(base$med_income < 60000)
tabyl(base$med_income < 120000)
tabyl(base$med_income < 100000)

## 60000 and 100000 works

```

Define categories for bivariate map
```{r}

base <- base %>%
  mutate(ci_base_change_cat = case_when(pct_change_ci_base < 0 ~ "C",
                                        pct_change_ci_base >= 0 & pct_change_ci_base <= 100 ~ "B",
                                        pct_change_ci_base > 100 ~ "A",
                                        TRUE ~ NA_character_),
         med_income_cat = case_when(med_income < 60000 ~ "1",
                                    med_income >= 60000 & med_income < 100000 ~ "2",
                                    med_income >= 100000 ~ "3",
                                    TRUE ~ NA_character_),
         bivar_ci_medinc = ifelse(!is.na(med_income_cat) & 
                                  !is.na(ci_base_change_cat) &
                                   excludedfrom_fd_yes_no == "N", paste0(ci_base_change_cat, med_income_cat), NA))

## Check
tabyl(base$ci_base_change_cat)
tabyl(base$med_income_cat)
tabyl(base$bivar_ci_medinc)

```


Define palettes

Diverging palette for CI contribution; need to custom make scale because the values are skewed (can't go lower than 100 in negative direction; going well over 100 in positive numbers) and R will set neutral color at midpoint of range, not at 0.

```{r}

## create special column to flag anyone whose change in ci tax base pct was > 500%, easier to map
base$pct_change_ci_simpl = if_else(base$pct_change_ci_base > 500, 500, base$pct_change_ci_base)

ci_bin_neg <- seq(-100, -0.01, by = 20)
ci_bin_pos <- seq(.01, 501, by = 100)
ci_bin_breaks <- c(ci_bin_neg, 0, ci_bin_pos)

ci_bin_names <- c("-100% to -80%", "-80% to -60%","-60% to -40%", "-40% to -20%", "-20% to 0%", "0%", "0% to 100%", "100% to 200%", "200% to 300%", "300% to 400%", "400%+")

base$ci_pct_change_bin <- cut(base$pct_change_ci_simpl, breaks = ci_bin_breaks)

levels(base$ci_pct_change_bin) <- ci_bin_names

```

save out
```{r}

saveRDS(base, file = "C:/Users/mmalloy/Documents/docs/fiscal_disparities_formatted.rds")

```

